The CrypTool Script: \\ Cryptography, Mathematics and more \\ \vspace{30pt} {\large Background reading\\*[-8pt] for the free eLearning program CrypTool (version 1.4.00)} \vspace{60pt}

### 2007-04-23 script/en/chapters/primes.tex
be changes dated 2006-04-09 2006-04-11 2006-05-05 2006-05-30 2006-07-26

\newpage
\section{Prime Numbers}
\hypertarget{Kapitel_2}{}
\label{Label_Kapitel_2}
(Bernhard Esslinger, May 1999, Updates Nov. 2000, Dec. 2001, June 2003, May 2005)

@@

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid%
\footnote{Euclid,
a Greek mathematician of 4th and 3rd century B.C. He worked at the
Egyptian academy of Alexandria and wrote ``The Elements'', the most well 
known systematically textbook of the Greek mathematics.} 
knew this in the third century B.C.
\begin{theorem}[Euclid\footnote{The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. The theorem has already been distinguished
and proven in Euclid's Elements (Book IX, theorem 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as followed
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than 
any previously existing amount of prime numbers.
}]\label{thm-pz-euklid} % Ende der Fußnote
% Fussnote VOR "]" in [Euclid], damit kein Leerraum vor der Fussnotennummer.
% Vorher stand da: ...(Euclid). BLANK Fussnote und das Blank stoerte.
% Nun steht die Fussnote direkt hinter "Euclid" und vor der ")".
% Eigentlich hätte ich sie gerne direkt hinter "(Euclid)", noch vor dem 
% automatisch gesetzten Punkt. 
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}
His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (proof by contradiction \index{Proof by contradiction}). 
He assumed that there is only a finite number of primes and therefore a 
largest prime number. Based on this assumption, he drew logical conclusions
until he obtained an obvious contradiction. This meant that something must
be wrong. As there were no mistakes in the chain of conclusions, it could
only be the assumption that was wrong. Therefore, there must be an infinite
number of primes!

@@

% --------------------------------------------------------------------------
\hypertarget{RecordPrimes}{}
\subsubsection{The 10 largest known primes (as of May 2005)} 
\index{Prime number!records}
\label{RecordPrimes}

The following table contains the "top ten" record primes and
a description of its particular number type\footnote{%
An up-to-date version can be found in the internet at
     \href{http://primes.utm.edu/largest.html}
  {\texttt{http://primes.utm.edu/largest.html}}.
}:

\begin{table}[h]
\begin{center}
\begin{tabular}{|c|cccc|}
\hline
	& {\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Description} \\
\hline
	1  & $2^{25,964,951}-1$ & 7,816,230 & 2005 & Mersenne, 42th known \\
	2  & $2^{24,036,583}-1$ & 7,235,733 & 2004 & Mersenne, 41th known \\
	3  & $2^{20,996,011}-1$ & 6,320,430 & 2003 & Mersenne, 40th known \\
	4  & $2^{13,466,917}-1$ & 4,053,946 & 2001 & Mersenne, M-39 \\
	5  & $28,433 \cdot 2^{7,830,457}+1$ & 2,357,207 & 2004 & Generalized Mersenne \\
	6  & $2^{ 6,972,593}-1$ & 2,098,960 & 1999 & Mersenne, M-38 \\
	7  & $5,359 \cdot 2^{5,054,502}+1$ & 1,521,561 & 2003 & Generalized Mersenne \\
	8  & $2^{ 3,021,377}-1$ &   909,526 & 2001 & Mersenne, M-37 \\
	9  & $2^{ 2,976,221}-1$ &   895,932 & 2001 & Mersenne, M-36 \\
	10 & $1,372,930^{131,072}+1$ &   804,474 & 2003 & Generalized Fermat\footnotemark \\

\hline
\end{tabular}
\caption{The 10 largest known primes and its particular number types (as of May 2005)}
\end{center}
\end{table} 
\footnotetext{$ 1,372,930^{131,072} + 1 = 1,372,930^{(2^{17})}+1 $ } 
%be_2005: Erzwingen, dass die Abb. noch in diesem Kapitel !

The largest known prime is a Mersenne prime.
This prime was found by the \hyperlink{GIMPS-project}{GIMPS project}.

@@

% --------------------------------------------------------------------------
\subsubsection{Special number types -- Mersenne numbers and Mersenne primes} 
\hypertarget{MersenneNumbers01}{}
\label{zahlentyp_mersenne}
\index{Mersenne!number}

@@

Mersenne claimed that $2^{67}-1$ is a prime number. There is also a mathematical
history behind this claim: it first took over 200 years before 
\index{Lucas, Edouard} Edouard Lucas (1842-1891) proved that this number 
is composite.
However, he argued indirectly and did not name any of the factors. Then Frank
Nelson Cole\index{Cole, Frank Nelson}\footnote{%
Frank Nelson Cole, American mathematician, Sep. 20, 1861 $-$ May 26, 1926.}
showed in 1903 which factors make up this composite number: 
$$ 2^{67} -1
=147, 573, 952, 589, 676, 412, 927 = 193, 707, 721 \cdot 761, 838, 257, 287. $$
He admitted to having worked 20 years on the factorisation 
\index{Factorisation} (expression as a product of prime factors)\footnote{%
  Using CrypTool\index{CrypTool} you can factorize numbers in the 
  following way: menu {\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem 
  \textbackslash{} Factorisation of a Number}. \\
  CrypTool can factorize in a reasonable time numbers no longer than 250 bit.
  Numbers bigger than 1024 bits are currently not accepted by CrypTool. \\
  The current factorization records are listed in chapter \ref{NoteFactorisation}.
  \index{Factorisation!factorisation records}
}
of this 21-digit decimal number!

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers 
\index{Mersenne!prime number} have the exponents\footnote{%
The following page from Landon Curt Noll\index{Noll, Landon Curt} contains 
all Mersenne primes including its date of discovery and its value as number 
and as word:
      \href{http://www.isthe.com/chongo/tech/math/prime/mersenne.html}
   {\texttt{http://www.isthe.com/chongo/tech/math/prime/mersenne.html}}. \\
Also see:
      \href{http://www.utm.edu/}
   {\texttt{http://www.utm.edu/}}.
                             } 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217; ~ 4,253;\\
 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; ~ 132,049; \\
 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221;
 ~ 3,021,377; ~ 6,972,593; \\
 ~ 13,466,917;  ~ 20,996,011;  ~ 24,036,583;  ~ 25,964,951.
% be_2005_UPDATEN_if-new-mersenne-prime-appears          ~ xxx,xxx,xxx.
\end{array}
$$
Thus
$42$            % be_2005_UPDATEN_if-new-mersenne-prime-appears
Mersenne prime numbers are currently known%
\index{Prime number!Mersenne}\index{Mersenne!prime number}. 
For the first 39            % be_2005_UPDATEN_if-new-mersenne-prime-appears
Mersenne prime numbers we know that this list is complete.
The exponents until the 40th     % be_2005_UPDATEN_if-new-mersenne-prime-appears
Mersenne prime number have not yet been checked completely\footnote{%
The current status of the check can be found at:
      \href{http://www.mersenne.org/status.htm}
   {\texttt{http://www.mersenne.org/status.htm}}.\\
Hints, how the primality of a number can be checked, are in chapter
\ref{primality_tests}, prime number tests\index{Prime number!test}.

@@

\vskip +25 pt
\paragraph{M-39 -- December 2001}%
\index{Mersenne!prime number!M-39}\mbox{}

@@

Right now (May 2005) all prime exponents smaller than $ 13.466.917 $ have been
tested and double-checked (see home page of the GIMPS project:
{\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}): so we can
be certain, that this is really the 39th Mersenne prime number and that
there are no smaller undiscovered Mersenne primes.

@@

Right now (as of May 2005) already 42 Mersenne primes are known.

@@

Right now the GIMPS project has discovered eight
            % be_2005_UPDATEN_if-new-mersenne-prime-appears
largest Mersenne primes so far, including the largest known prime number at all.

The following table contains these Mersenne record primes\footnote{%
An up-to-date version can be found in the internet at
     \href{http://www.mersenne.org/history.htm}
  {\texttt{http://www.mersenne.org/history.htm}}.
}:

            % be_2005_UPDATEN_if-new-mersenne-prime-appears
\begin{table}[h]
\begin{center}
\begin{tabular}{|cccc|}
\hline
	{\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Who} \\
\hline
	$2^{25,964,951}-1$ & 7,816,230 & February 18, 2005 & Dr. Martin Nowak \\
	$2^{24,036,583}-1$ & 7,235,733 & May 15, 2004      & Josh Findley     \\
	$2^{20,996,011}-1$ & 6,320,430 & November 17, 2003 & Michael Shafer   \\
	$2^{13,466,917}-1$ & 4,053,946 & November 14, 2001 & Michael Cameron  \\
	$2^{ 6,972,593}-1$ & 2,098,960 & June 1, 1999      & Nayan Hajratwala \\
	$2^{ 3,021,377}-1$ &   909,526 & January 27, 1998  & Roland Clarkson  \\
	$2^{ 2,976,221}-1$ &   895,932 & August 24, 1997   & Gordon Spence    \\
	$2^{ 1,398,269}-1$ &   420,921 & November 1996     & Joel Armengaud   \\

\hline
\end{tabular}
\caption{The largest primes found by the GIMPS project (as of May 2005)}
\end{center}
\end{table}

@@

If you are not used to calculating with remainders (modulo), please simply
accept the theorem or first read \hyperlink{Kapitel_3}
{chapter \ref{Kapitel_3} ``Introduction to Elementary Number Theory with 
Examples''}.  What is important here is that this sentence implies that if
this equation is not met for any whole number $a$, then $p$ is not a prime! The
tests (e.g. for the first formulation) can easily be performed using the {\em
test basis} $a = 2$.

@@

\vskip +25 pt
\paragraph{Carmichael numbers}%
\index{Number!Carmichael}%
\hypertarget{HT-Carmichael-number01}{}\label{L-Carmichael-number01}%
\mbox{}
\vskip +10 pt
There are numbers that pass the Fermat test with all bases and yet are not
prime: these numbers are called {\em Carmichael numbers}. The first of these is
$$ 561 = 3 \cdot 11 \cdot 17 .$$

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\subsubsection{Mersenne numbers $f(n) = 2^n - 1$  \quad for $ n $ prime}
    \hypertarget{MersenneNumbers02}{}
    \index{Prime number!Mersenne} \index{Mersenne!prime number}
    As shown \hyperlink{MersenneNumbers01}{above}, this formula seems
    to deliver relatively large prime numbers but - as for $n=11$ 
    [$f(n)=2,047$] - it is repeatedly the case that the result even with
    prime exponents is \hyperlink{Mer-nums-not-always-prim}{not} prime.\\
    Today, all the Mersenne primes having less than around 4,000,000 digits are 
    known (M-39\index{Mersenne!prime number!M-39}):
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}
\end{itemize}
            % be_2005_UPDATEN_if-new-mersenne-prime-appears

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\subsubsection
   [Generalized Mersenne numbers $ f(b,n) = b^n \pm 1 $ / Cunningham project]
   {Generalized Mersenne numbers 
   $ f(b,n) = b^n \pm 1 $ / The Cunningham project}
   \hypertarget{generalizedMersennenumbers}{}

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\subsubsection[Fermat numbers $f(n) = 2^{2^n} + 1$]
              {Fermat numbers\footnotemark~$f(n) = 2^{2^n} + 1$} 
    \footnotetext{%
       The Fermat prime numbers play a role in circle division.
       As proven by Gauss\index{Gauss, Carl Friedrich} a regular $p$-edge
       can only be constructed with the use of a pair of compasses and a 
       ruler, when $p$ is a Fermat prime number.
    }
    \hypertarget{FermatNumbers02}{}
    \label{L-FermatNumbers02}
    \index{Prime number!Fermat} \index{Fermat!prime number}       
    As mentioned \hyperlink{FermatNumbers01}{above} in 
    chapter~\ref{FermatNumbers01}, Fermat wrote to Mersenne regarding 
    his assumption, that all numbers of this type are primes.
    This assumption was disproved by Euler (1732). The prime $641$ divides $f(5)$\footnote{%
    Surprisingly this number can easily be found by using Fermat's theorem (see e.g. \cite[p. 176]{Scheid1994})
    }. 
%    Surprisingly he would have been able obtain a positive result using
%    the negative prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
f(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
f(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
f(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
f(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
f(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = \mbox{65,537} &  \mapsto {\rm ~prime}  \\
f(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = \mbox{4,294,967,297} = 641 \cdot \mbox{6,700,417} &  \mapsto {\rm ~NOT~prime} ! \\
f(6) = 2^{2^6} + 1  = 2^{64} + 1 &  = \mbox{18,446,744,073,709,551,617} \\
                                 &  = \mbox{274,177} \cdot \mbox{67,280,421,310,721} & \mapsto {\rm ~NOT~prime} !\\
f(7) = 2^{2^7} + 1  = 2^{128} + 1 & = \mbox{(see page~\pageref{F7Morrison})}  
																 & \mapsto {\rm ~NOT~prime} !

@@

%    ({\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
%           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}).
% M.E. ist die Zahl auf seiner Webseite zu groß, da M-39 "nur" 4 Mio. Stellen hat !
%    Heute kennt man alle Fermatschen Primzahlen mit bis zu 2.000.000.000 ??? 
%    Dezimalstellen. \\

% --------------------------------------------------------------------------
\vskip +10 pt
\subsubsection[Generalized Fermat numbers $f(b,n) = b^{2^n} + 1$]
              {Generalized Fermat numbers\footnotemark~$f(b,n) = b^{2^n} + 1$}
    \footnotetext{%
      The base of this power is no longer restricted to 2 . \\
      Even more generic would be:  $f(b,c,n) = b^{c^n} \pm 1$
    }
\hypertarget{generalizedFermatprimes}{}
\label{generalized-fermat}
\index{Fermat!number!generalized} 
    Generalized Fermat numbers are more numerous than Mersenne numbers of a
    equal size and many of them are waiting to be discovered to fill the
    big gaps between the Mersenne primes already found or still undiscovered.
    Progress in number theory made it possible that numbers, where the
    representation is not limited to the base 2, can be tested at almost
    the same speed than a Mersenne number.

>>>>

\newpage
\hypertarget{Kapitel_2}{}
\section{Prime Numbers}
\label{Label_Kapitel_2}
(Bernhard Esslinger, May 1999; Updates Nov. 2000, Dec. 2001, June 2003, May 2005, March 2006)

@@

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid%
\footnote{Euclid,
a Greek mathematician of 4th and 3rd century B.C. He worked at the
Egyptian academy of Alexandria and wrote ``The Elements'', the most well 
known systematically textbook of the Greek mathematics.} 
knew this in the third century B.C.
\begin{theorem}[Euclid\footnote{The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. The theorem has already been distinguished
and proven in Euclid's Elements (Book IX, theorem 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as followed
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than 
any previously existing amount of prime numbers.
}]\label{thm-pz-euklid} % Ende der Fuﬂnote
% Fussnote VOR "]" in [Euclid], damit kein Leerraum vor der Fussnotennummer.
% Vorher stand da: ...(Euclid). BLANK Fussnote und das Blank stoerte.
% Nun steht die Fussnote direkt hinter "Euclid" und vor der ")".
% Eigentlich h‰tte ich sie gerne direkt hinter "(Euclid)", noch vor dem 
% automatisch gesetzten Punkt. 
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}
His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (proof by contradiction \index{Proof by contradiction}). 
He assumed that there is only a finite number of primes and therefore a 
largest prime number. Based on this assumption, he drew logical conclusions
until he obtained an obvious contradiction. This meant that something must
be wrong. As there were no mistakes in the chain of conclusions, it could
only be the assumption that was wrong. Therefore, there must be an infinite
number of primes!

@@

% --------------------------------------------------------------------------
\hypertarget{RecordPrimes}{}
\subsubsection{The 11 largest known primes (as of March 2006)} 
\label{RecordPrimes}
\index{Prime number!records}

The following table contains the current record primes and
a description of its particular number type\footnote{%
An up-to-date version can be found in the internet at
     \href{http://primes.utm.edu/largest.html}
  {\texttt{http://primes.utm.edu/largest.html}}.
}:

\begin{table}[h]
\begin{center}
\begin{tabular}{|c|cccc|}
\hline    % Eyecatcher_neue-Mersenne
	& {\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Description} \\
\hline
	1  & $2^{30,402,457}-1$ & 9,152,052 & 2005 & Mersenne, 43rd known \\
	2  & $2^{25,964,951}-1$ & 7,816,230 & 2005 & Mersenne, 42nd known \\
	3  & $2^{24,036,583}-1$ & 7,235,733 & 2004 & Mersenne, 41st known \\
	4  & $2^{20,996,011}-1$ & 6,320,430 & 2003 & Mersenne, 40th known \\
	5  & $2^{13,466,917}-1$ & 4,053,946 & 2001 & Mersenne, M-39 \\
	6  & $28,433 \cdot 2^{7,830,457}+1$ & 2,357,207 & 2004 & Generalized Mersenne \\
	7  & $2^{ 6,972,593}-1$ & 2,098,960 & 1999 & Mersenne, M-38 \\
	8  & $5,359 \cdot 2^{5,054,502}+1$ & 1,521,561 & 2003 & Generalized Mersenne \\
	9  & $2^{ 3,021,377}-1$ &   909,526 & 2001 & Mersenne, M-37 \\
	10  & $2^{ 2,976,221}-1$ &   895,932 & 2001 & Mersenne, M-36 \\
	11 & $1,372,930^{131,072}+1$ &   804,474 & 2003 & Generalized Fermat\footnotemark \\

\hline
\end{tabular}
\caption{The 11 largest known primes and its particular number types
         (as of March 2006)}
\end{center}
\end{table} 
\footnotetext{$ 1,372,930^{131,072} + 1 = 1,372,930^{(2^{17})}+1 $ } 
%be_2005: Erzwingen, dass die Abb. noch in diesem Kapitel !

The largest known prime is a Mersenne prime.
This prime was found by the \hyperlink{GIMPS-project}{GIMPS project}
(chapter~\ref{zahlentyp_mersenne}).

@@

% --------------------------------------------------------------------------
\hypertarget{MersenneNumbers01}{}
\subsubsection{Special number types -- Mersenne numbers and Mersenne primes} 
\label{zahlentyp_mersenne}
\index{Mersenne!number}

@@

Mersenne claimed that $2^{67}-1$ is a prime number. There is also a mathematical
history behind this claim: it first took over 200 years before 
\index{Lucas, Edouard} Edouard Lucas (1842-1891) proved that this number 
is composite.
However, he argued indirectly and did not name any of the factors. Then Frank
Nelson Cole\index{Cole, Frank Nelson}\footnote{%
Frank Nelson Cole, American mathematician, Sep. 20, 1861 $-$ May 26, 1926.}
showed in 1903 which factors make up this composite number: 
$$ 2^{67} -1
=147, 573, 952, 589, 676, 412, 927 = 193, 707, 721 \cdot 761, 838, 257, 287. $$
He admitted to having worked 20 years on the factorisation 
\index{Factorisation} (expression as a product of prime factors)\footnote{%
  Using CrypTool\index{CrypTool} you can factorize numbers in the 
  following way: menu {\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem 
  \textbackslash{} Factorisation of a Number}. \\
  CrypTool can factorize in a reasonable time numbers no longer than 250 bit.
  Numbers bigger than 1024 bits are currently not accepted by CrypTool. \\
  The current factorization records are listed in chapter \ref{NoteFactorisation}.
  \index{Factorisation!factoring records}
}
of this 21-digit decimal number!

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers 
\index{Mersenne!prime number} have the exponents\footnote{%
The following page from Landon Curt Noll\index{Noll, Landon Curt} contains 
all Mersenne primes including its date of discovery and its value as number 
and as word:
      \href{http://www.isthe.com/chongo/tech/math/prime/mersenne.html}
   {\texttt{http://www.isthe.com/chongo/tech/math/prime/mersenne.html}}. \\
Also see:
      \href{http://www.utm.edu/}
   {\texttt{http://www.utm.edu/}}.
                             } 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217; ~ 4,253;\\
 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; ~ 132,049; \\
 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221;
 ~ 3,021,377; ~ 6,972,593; \\
 ~ 13,466,917;  ~ 20,996,011;  ~ 24,036,583;  ~ 25,964,951;  ~ 30,402,457.
% be_2005_UPDATEN_if-new-mersenne-prime-appears          ~ xxx,xxx,xxx.
\end{array}
$$
Thus    % Eyecatcher_neue-Mersenne
$43$            % be_2005_UPDATEN_if-new-mersenne-prime-appears
Mersenne prime numbers are currently known%
\index{Prime number!Mersenne}\index{Mersenne!prime number}. 
For the first 39            % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
Mersenne prime numbers we know that this list is complete.
The exponents until the 40th     % be_2005_UPDATEN_if-new-mersenne-prime-appear    % Eyecatcher_neue-Mersennes
Mersenne prime number have not yet been checked completely\footnote{%
The current status of the check can be found at:
      \href{http://www.mersenne.org/status.htm}
   {\texttt{http://www.mersenne.org/status.htm}}.\\
Hints, how the primality of a number can be checked, are in chapter
\ref{primality_tests}, prime number tests\index{Prime number!test}.

@@

\vskip +25 pt
\hypertarget{M-39}{}
\paragraph{M-39 -- December 2001}%
\index{Mersenne!prime number!M-39}\mbox{}

@@

Right now (May 2005) all prime exponents smaller than $ 13.466.917 $ have been
tested and double-checked (see home page of the GIMPS project:
{\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}): so we can
be certain, that this is really the 39th Mersenne prime number and that
there are no smaller undiscovered Mersenne primes (it is common usage to use
the notation M-nn not until it is proven, that the nn-th known Mersenne prime
is really the nn-th Mersenne prime).

@@

@@

Right now the GIMPS project has discovered nine
   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
largest Mersenne primes so far, including the largest known prime number at all.

The following table contains these Mersenne record primes\footnote{%
An up-to-date version can be found in the internet at
     \href{http://www.mersenne.org/history.htm}
  {\texttt{http://www.mersenne.org/history.htm}}.
}$^,$\footnote{%
Always, when a new record is published in the respective forums the same
and often ironic discussions start: Is there a deeper sense? Can this result
be applied for anything useful?
The answer is, that we don't know it yet. In fundamental research one
cannot see at once how it brings mankind forward.
}:

   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
\begin{table}[h]
\begin{center}
\begin{tabular}{|cccc|}
\hline
	{\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Who} \\
\hline
	$2^{30,402,457}-1$ & 9,152,052 & December 15, 2005 & Curtis Cooper/Steven Boone \\
	$2^{25,964,951}-1$ & 7,816,230 & February 18, 2005 & Martin Nowak     \\
	$2^{24,036,583}-1$ & 7,235,733 & May 15, 2004      & Josh Findley     \\
	$2^{20,996,011}-1$ & 6,320,430 & November 17, 2003 & Michael Shafer   \\
	$2^{13,466,917}-1$ & 4,053,946 & November 14, 2001 & Michael Cameron  \\
	$2^{ 6,972,593}-1$ & 2,098,960 & June 1, 1999      & Nayan Hajratwala \\
	$2^{ 3,021,377}-1$ &   909,526 & January 27, 1998  & Roland Clarkson  \\
	$2^{ 2,976,221}-1$ &   895,932 & August 24, 1997   & Gordon Spence    \\
	$2^{ 1,398,269}-1$ &   420,921 & November 1996     & Joel Armengaud   \\

\hline
\end{tabular}
\caption{The largest primes found by the GIMPS project (as of March 2006)}
\end{center}
\end{table}

@@

If you are not used to calculate with remainders (modulo), please simply
accept the theorem or first read \hyperlink{Chapter_ElementaryNT}
{chapter \ref{Chapter_ElementaryNT} ``Introduction to Elementary Number Theory with 
Examples''}.  What is important here is that this sentence implies that if
this equation is not met for any whole number $a$, then $p$ is not a prime! The
tests (e.g. for the first formulation) can easily be performed using the {\em
test basis} $a = 2$.

@@

\vskip +25 pt
\paragraph{Carmichael numbers}%
\index{Number!Carmichael}%
\hypertarget{HT-Carmichael-number01}{}\label{L-Carmichael-number01}%
\mbox{}
\vskip +10 pt
There are pseudo prime numbers n that pass the Fermat test 
$$ a^{n-1} \equiv 1 \;{\rm mod}\; n $$
with all bases a which are relatively prime to n [$ gcd (a,n) = 1 $],
despite these numbers n are not prime: These numbers are called
{\em Carmichael numbers}.
The first of these is $$ 561 = 3 \cdot 11 \cdot 17 .$$

Sample: The number to be tested is 561. Because $ 561 = 3 \cdot 11 \cdot 17 $
it is: \\
The test condition $ a^{560} \;{\rm mod}\; 561 = 1 $ is satified 
for $a = 2, 4, 5, 7, \cdots $,\\
but not for $a = 3, 6, 9, 11, 12, 15, 17, 18, 21, 22, \cdots $.\\ 
This means the test condition must not be satisfied for multiples of the
prime factors 3, 11 or 17.\\
The test applied for $a=3$ results in: $ 3^{560} \;{\rm mod}\; 561 = 375 $.\\
The test applied for $a=5$ results in: $ 5^{560} \;{\rm mod}\; 561 = 1 $.

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{MersenneNumbers02}{}
\subsubsection{Mersenne numbers $f(n) = 2^n - 1$  \quad for $ n $ prime}
    \index{Prime number!Mersenne} \index{Mersenne!prime number}
    As shown \hyperlink{MersenneNumbers01}{above}, this formula seems
    to deliver relatively large prime numbers but - as for $n=11$ 
    [$f(n)=2,047$] - it is repeatedly the case that the result even with
    prime exponents is \hyperlink{Mer-nums-not-always-prim}{not} prime.\\
    Today, all the Mersenne primes having less than around 4,000,000 digits
    are known (\hyperlink{M-39}{M-39}\index{Mersenne!prime number!M-39}):
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}
\end{itemize}      % be_2005_UPDATEN_if-new-mersenne-prime-appears

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{generalizedMersennenumbers}{}
\subsubsection
   [Generalized Mersenne numbers $ f(b,n) = b^n \pm 1 $ / Cunningham project]
   {Generalized Mersenne numbers 
   $ f(b,n) = b^n \pm 1 $ / The Cunningham project}

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{FermatNumbers02}{}
\subsubsection[Fermat numbers $f(n) = 2^{2^n} + 1$]
              {Fermat numbers\footnotemark~$f(n) = 2^{2^n} + 1$} 
    \footnotetext{%
       The Fermat prime numbers play a role in circle division.
       As proven by Gauss\index{Gauss, Carl Friedrich} a regular $p$-edge
       can only be constructed with the use of a pair of compasses and a 
       ruler, when $p$ is a Fermat prime number.
    }
    \label{L-FermatNumbers02}
    \index{Prime number!Fermat} \index{Fermat!prime number}       
    As mentioned \hyperlink{FermatNumbers01}{above} in 
    chapter~\ref{FermatNumbers01}, Fermat wrote to Mersenne regarding 
    his assumption, that all numbers of this type are primes.
    This assumption was disproved by Euler (1732). The prime $641$ divides $f(5)$\footnote{%
    Surprisingly this number can easily be found by using Fermat's theorem (see e.g. \cite[p. 176]{Scheid1994})
    }. 
%    Surprisingly he would have been able obtain a positive result using
%    the negative prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
f(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
f(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
f(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
f(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
f(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = \mbox{65,537} &  \mapsto {\rm ~prime}  \\
f(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = \mbox{4,294,967,297} = 641 \cdot \mbox{6,700,417} &  \mapsto {\rm ~NOT~prime} ! \\
f(6) = 2^{2^6} + 1  = 2^{64} + 1 &  = \mbox{18,446,744,073,709,551,617} \\
                                 &  = \mbox{274,177} \cdot \mbox{67,280,421,310,721} & \mapsto {\rm ~NOT~prime} !\\
f(7) = 2^{2^7} + 1  = 2^{128} + 1 & = \mbox{(see page~\pageref{F7Morrison})}  
																 & \mapsto {\rm ~NOT~prime} !

@@

%    ({\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
%           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}).
% M.E. ist die Zahl auf seiner Webseite zu groﬂ, da M-39 "nur" 4 Mio. Stellen hat !
%    Heute kennt man alle Fermatschen Primzahlen mit bis zu 2.000.000.000 ??? 
%    Dezimalstellen. \\

% --------------------------------------------------------------------------
\vskip +10 pt
\hypertarget{generalizedFermatprimes}{}
\subsubsection[Generalized Fermat numbers $f(b,n) = b^{2^n} + 1$]
              {Generalized Fermat numbers\footnotemark~$f(b,n) = b^{2^n} + 1$}
    \footnotetext{%
      The base of this power is no longer restricted to 2 . \\
      Even more generic would be:  $f(b,c,n) = b^{c^n} \pm 1$
    }
\label{generalized-fermat}
\index{Fermat!number!generalized} 
    Generalized Fermat numbers are more numerous than Mersenne numbers of a
    equal size and many of them are waiting to be discovered to fill the
    big gaps between the Mersenne primes already found or still undiscovered.
    Progress in number theory made it possible that numbers, where the
    representation is not limited to the base 2, can be tested at almost
    the same speed than a Mersenne number.
